Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

### 2025-03-17 code/processing-code/processing-code-tickcases.qmd
Part 3 Done
Made some better manuscript edits

Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time? One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.

>>>>

Another question that I want to analyze later and therefore needed to clean for was "what is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?" One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.

### 2025-03-17 products/manuscript/Manuscript.qmd
Part 3 Done
Made some better manuscript edits

Tick-borne illnesses are diseases that humans can contract from the bite
of an infected tick. Ticks are small, blood-sucking arachnids that are
often found in wooded, grassy, or suburban areas. When a tick feeds on
an infected animal, it can acquire pathogens—such as bacteria, viruses,
or parasites—that can then be transmitted to humans. Examples of
tick-bourne diseases include, but are not limited to, Lyme disease,
anaplasmosis, and babesiosis. These diseases are of increasing public
health concern, especially as the geographic range and seasonality of
ticks expand due to environmental and ecological changes.

@@

Looked for Missing Data & General Trends

-   Identified Disease Burden Trends & Hotspot Counties

<!-- -->

-   Generated a stacked bar graph to visualize case distribution across
    regions. - Checked Age Distribution

<!-- -->

-   Verified that no cases had missing age values in the cleaned
    dataset.

<!-- -->

-   Conducted an exploratory analysis of disease counts by age group.

-   Evaluated Data Completeness Over Time

-   Extracted onset date information to track missing data trends.

-   Created a plot distinguishing between blank and non-blank STATUS
    entries over time.

-   Identified periods of high missingness, which may indicate data
    entry errors or reporting gaps.

Analyzed Timeline Between Key Case Events

-   Calculated time differences for:

    -   Onset date (DOO) to first lab test date (LABDATE1)

    -   First lab test (LABDATE1) to second lab test (LABDATE2)

    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)

    -   Summed these values to determine the total case duration. -
        Saved the processed dataset as a CSV file for further
        analysis. - Created a formatted HTML table to present extracted
        time intervals.

Addressed Data Quality Issues

-   Identified negative time values in lab test dates, indicating data
    entry errors

-   Implemented a filtering step to remove records with invalid date
    sequences

@@

A grouped bar chart maps case counts across
Georgia counties, categorized by disease type. Due to the density of
information, this visualization is refined to focus on the top five
tick-borne diseases and their distribution across the 12 counties with
the highest confirmed case counts.
```{r}
library(here)
library(knitr)

@@

Age distribution is assessed as part of
the initial data cleaning process, confirming that no cases in the
cleaned dataset have missing age values. Additionally, an exploratory
analysis examines disease counts by age group to identify potential
patterns in disease prevalence.
```{r}
library(here)
library(knitr)

@@

**Data Completeness Over Time** 
To evaluate how data completeness has
evolved, the dataset is processed to extract onset date information. A
visualization distinguishes between blank and non-blank entries in the
STATUS column over time, helping to identify periods with high
missingness. Spikes in blank values may indicate data entry errors,
system issues, or reporting gaps that should be addressed before
conducting further analysis.

@@

**Onset date (DOO) to first lab test date (LABDATE1)**

**First lab test (LABDATE1) to second lab test (LABDATE2)**

**Second lab test (LABDATE2) to case update date (UPDATEDATE)**

@@

*Explain anything related to your statistical analyses.*

@@

>>>>

Tick-borne illnesses are diseases that humans can contract from the bite
of an infected tick. Ticks are small, blood-sucking arachnids that are
often found in wooded, grassy, or suburban areas. When a tick feeds on
an infected animal, it can acquire pathogens—such as bacteria, viruses,
or parasites—that can then be transmitted to humans. Examples of
tick-bourne diseases include, but are not limited to, Lyme disease,
Anaplasmosis, and babesiosis. These diseases are of increasing public
health concern, especially as the geographic range and seasonality of
ticks expand due to environmental and ecological changes.

@@

**Looked for Missing Data & General Trends**

-   Excluded cases with a missing or non-“CONFIRMED” STATUS to retain
    only confirmed cases.
-   Refined the dataset by retaining only confirmed “positive” test
    results, excluding missing or ambiguous results, leaving 95 valid
    cases.
-   Additional filtering ensured correct identification of positive
    cases while excluding non-positive values.

**Identified Disease Burden Trends & Hotspot Counties**

-   Generated a histogram to visualize the distribution of tick-borne
    diseases in Georgia between 2019 and 2024.
-   Created a bar chart highlighting the five most common diseases.
-   Specifically analyzed Lyme Disease cases, focusing on test results.

**Checked Age Distribution**

-   Verified that no cases had missing age values in the cleaned
    dataset.
-   Conducted an exploratory analysis of disease counts by age group.

**Evaluated Data Completeness Over Time**

-   Extracted onset date information to track missing data trends.
-   Created a plot distinguishing between blank and non-blank STATUS
    entries over time to identify periods of high missingness,
    indicating possible data entry errors or reporting gaps.

**Analyzed Timeline Between Key Case Events**

-   Calculated time differences for:
    -   Onset date (DOO) to first lab test date (LABDATE1)
    -   First lab test (LABDATE1) to second lab test (LABDATE2)
    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
-   Summed these values to determine the total case duration.
-   Saved the processed dataset as a CSV file for further analysis.
-   Created a formatted HTML table to present the extracted time
    intervals.

**Addressed Data Quality Issues**

-   Identified negative time values between lab test dates, indicating
    data entry errors.
-   Implemented a filtering step to remove records with invalid date
    sequences.

@@

A grouped bar chart maps case counts across Georgia counties,
categorized by disease type. Due to the density of information, this
visualization is refined to focus on the top five tick-borne diseases
and their distribution across the 12 counties with the highest confirmed
case counts.

@@

Age distribution is assessed as part of the initial data cleaning
process, confirming that no cases in the cleaned dataset have missing
age values. Additionally, an exploratory analysis examines disease
counts by age group to identify potential patterns in disease
prevalence.

@@

**Data Completeness Over Time**

To evaluate how data completeness has evolved, the dataset is processed
to extract onset date information. A visualization distinguishes between
blank and non-blank entries in the STATUS column over time, helping to
identify periods with high missingness. Spikes in blank values may
indicate data entry errors, system issues, or reporting gaps that should
be addressed before conducting further analysis.

@@

-   **Onset date (DOO) to first lab test date (LABDATE1)**

-   **First lab test (LABDATE1) to second lab test (LABDATE2)**

-   **Second lab test (LABDATE2) to case update date (UPDATEDATE)**

@@

@@

For the first key question related to the disease trends over time (in
years). In looking at the plot, it appears that there have been
increases over time were observed in most of the diseases. However, the
rate of increase was relatively slow as indicated by the lack of
statistical significance of these trends (p\>0.5) which may indicate an
attribution to the increase in reporting rather than an increase in
overall disease burden. In the full analysis and second iteration of
this, these trends will be compared year to year, comparing each year to
the 2018 standard.

+------------+------------+------------+------------+------------+
| term       | estimate   | std.error  | statistic  | p.value    |
|            |            |            |            |            |
| \<chr\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    |
+:===========+===========:+===========:+===========:+===========:+
| (          | -2         | 1          | -          | 0.1729390  |
| Intercept) | 3          | 573.052665 | 1.48025251 |            |
|            | 28.5151515 |            |            |            |
+------------+------------+------------+------------+------------+
| YEAR       | 1.1515152  | 0.777581   | 1.48089418 | 0.1727716  |
+------------+------------+------------+------------+------------+
| DISEAS     | 0.4242424  | 5.484565   | 0.07735207 | 0.9400358  |
| E          |            |            |            |            |
| BABESIOSIS |            |            |            |            |
+------------+------------+------------+------------+------------+
| DI         | 4.8484848  | 6.364775   | 0.76176850 | 0.4656907  |
| S          |            |            |            |            |
| EASEDENGUE |            |            |            |            |
+------------+------------+------------+------------+------------+
| DISEA      | 5.0000000  | 5.157889   | 0.96938889 | 0.3576794  |
| S          |            |            |            |            |
| EEHRLICHIA |            |            |            |            |
| C          |            |            |            |            |
| HAFFEENSIS |            |            |            |            |
+------------+------------+------------+------------+------------+
| DISEA      | -1.1515152 | 6.364775   | -          | 0.8604390  |
| S          |            |            | 0.18092002 |            |
| EEHRLICHIA |            |            |            |            |
| EWINGII    |            |            |            |            |
+------------+------------+------------+------------+------------+
| D          | 8.5606061  | 4.963745   | 1.72462659 | 0.1186814  |
| ISEASELYME |            |            |            |            |
| DISEASE    |            |            |            |            |
+------------+------------+------------+------------+------------+
| D          | 0.5757576  | 5.484565   | 0.10497781 | 0.9186958  |
| I          |            |            |            |            |
| SEASEROCKY |            |            |            |            |
| MOUNTAIN   |            |            |            |            |
| SPOTTED    |            |            |            |            |
| FEVER      |            |            |            |            |
+------------+------------+------------+------------+------------+
| D          | 8.2272727  | 5.593721   | 1.47080497 | 0.1754208  |
| ISEASEWEST |            |            |            |            |
| NILE (WNV) |            |            |            |            |
| INFECTION  |            |            |            |            |
+------------+------------+------------+------------+------------+

9 rows

# Generate correct file path
img_path <- here::here("results", "figures", "trend_over_time_by_disease.png")

The most interesting finding from the initial modeling was in relation
to the data completeness research question. In modeling the differences
across time in whether a final STATUS was inputted rather than left
blank, a general decrease was seen across the full 2018-2024 time scale.
Specifically, completion rates of the reporting form were approximately
50% in 2018 but dropped rapidly to 20% in 2019-- around which the rate
of completeness has hovered since 2019. However, in linear modeling, the
changes across time in the aggregate, p \> 0.05.

# Generate correct file path
img_path <- here::here("results", "figures", "completeness_over_time.png")

+------------+------------+------------+------------+------------+
| term       | estimate   | std.error  | statistic  | p.value    |
|            |            |            |            |            |
| \<chr\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    |
+:===========+===========:+===========:+===========:+===========:+
| (          | 9          | 5          | 1.733920   | 0.1434648  |
| Intercept) | 935.267857 | 729.946482 |            |            |
+------------+------------+------------+------------+------------+
| YEAR       | -4.910714  | 2.835202   | -1.732051  | 0.1438108  |
+------------+------------+------------+------------+------------+

2 rows

On a positive note for the utilization of the reporting system and
trends towards better tick-bourne disease response, there has been
statistically significant decreases in the time between oneset and an
initial lab visit/interaction. In 2018, the average time between these
two occurrences ranged from approximately 150-520 days. With marked
decreases, the 2024 data indicates that patients with tick-bourne
disease occurrences may be interacting with providers or lab testing in
the same month, same week or even same day.

# Generate correct file path
img_path <- here::here("results", "figures", "days_to_lab_over_time.png")
